---
title: "DSC 520 Week 03 - Descriptive Statistics and Plots"
author: "Daniel Adair"
date: "2026-09-28"
output: html_document
---
# Instructions 

Taking the same data set from Week 2, use R to create all of the descriptive statistical values and plots you learned this week. This week, R Markdown should be used to explain the statistical meaning and importance of what you’re doing; be critical, curious, and opinionated.

Specifically, for each numeric attribute in the dataset:

* Show the distributions of values visually. Explain, in comments, what those distributions mean statistically. Be thorough.
* Show the values of central tendency, both numerically and visually. Interpret those values in your own words in Markdown.
* Show the measures of dispersion, visually and numerically (where appropriate). Interpret those values in Markdown.
* Visually depict the Interquartile range. Interpret the IQR for the dataset (in Markdown).
* Create the statistics moments for the dataset.
* Experiment with histogram binning for the categorical data in R. This is open-ended intentionally because this dataset isn’t ideal for binning, but I want you to build something using binning just to see how it works 

--- 

#### Initial Setup 

Tidyverse is a pre-requisite library, and the initial import captures the sales data sample. 

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse) 
library(dplyr)

sales_raw <- read_csv("~/dev/msds/dsc520/sales_data_sample.csv") 

sales_cleaned <- sales_raw %>% mutate(
  Product = as.factor(Product),  
  Region = as.factor(Region),  
  OrderDate = as.factor(as.Date(OrderDate, format = "%m/%d/%Y"))
  ) 

# sales_summarized <- sales_cleaned %>% 
#   group_by(Product, Region) %>% 
#   summarise(
#     Sales_Sum = sum(TotalSales, na.rm = TRUE), 
#     Sales_Mean = mean(TotalSales, na.rm = TRUE), 
#     Order_Count = n(), 
#     Quantity_Sum = sum(Quantity, na.rm = TRUE), 
#     .groups = "drop" 
#     ) 

view(sales_cleaned)
glimpse(sales_cleaned)
```

--- 

# 1 - Measures of Central Tendency 

## 1.1 - Mean 

```{r 1-1 Mean} 

mean_sales <- mean(sales_cleaned$TotalSales)
paste0("Mean Total Sales: $", round(mean_sales, 2)) 

ggplot(sales_cleaned, aes(x = sales_cleaned$TotalSales)) + 
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
    labs(title = "Total Sales Distribution", x = "Total Sales", y = "Frequency") +
    theme_minimal()
       
hist(sales_cleaned$TotalSales) 

# Calculate the mean
# mean_val <- mean(data$rainfall, na.rm = TRUE)

# Render the distribution with the mean overlaid
ggplot(sales_cleaned, aes(x = TotalSales)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_sales), 
             color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = mean_sales, y = Inf, label = paste("Mean:", round(mean_sales, 2)), 
           vjust = 2, hjust = -0.1, color = "red", fontface = "bold") +
  labs(
    title = "Total Sales Distribution with Mean",
    x = "Total Sales",
    y = "Frequency"
  ) +
  theme_minimal()


paste0("Mean Unit Price: $", round(mean(sales_cleaned$UnitPrice), 2)) 
paste0("Mean Quantity: ", round(mean(sales_cleaned$Quantity), 2)) 

```



## 1.2 - Median 
## 1.3 - Mode 

--- 

# 2 - Measures of Dispersion 

## 2.1 - Variance 
## 2.2 - Standard Deviation 

sd(sales_cleaned$TotalSales)


## 2.3 - Heteroscedasticity & Homooscedasticity 
## 2.4 - Full Width at Half Maximum (FWHM) 
## 2.5 - Fano Factor and Coefficient of Variation 
## 2.6 - Interquartile Range 
## 2.7 - QQ Plots 

--- 

# 3 - Statistical Moments 

--- 

# 4 - Histograms 


--- 
