---
title: "dsc520-week02-adair-d"
author: "Daniel Adair"
date: "2026-09-15"
output: html_document
---
This is a test file for graphing. 

```{r}
# x <- c(11, 33, 54, 86, 211, 4)
x <- rnorm(50) 
x
summary(x)
hist(x) 
boxplot(x)
```

--- 

## 2.2 Exercise 

Taking this dataset, create the following plots in R:

1. Bar graph showing product and the sum of sales for each.
2. Pie chart showing sum of sales by region.
3. Box plots for average sales by region and average quantities by region.
4. Histogram of the count of orders by product.
5. A radial plot showing total sales (for all products) by month. 


--- 

### Data Processing 

Tidyverse is a pre-requisite library 

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

The following steps: 
- import the CSV (sales_raw) 
- transfer the raw data for categorization and data formatting via as.factor and mdy() 
- transfer the cleaned data into sales_summarized 
- summarize the cleaned data 

```{r}
sales_raw <- read_csv("~/dev/msds/dsc520/sales_data_sample.csv")

sales_cleaned <- sales_raw %>% mutate(
  Product = as.factor(Product),  
  Region = as.factor(Region),  
  OrderDate = as.Date(OrderDate, format = "%m/%d/%Y")
  ) 

sales_summarized <- sales_cleaned %>% 
  group_by(Product, Region) %>% 
  summarise(
    Sales_Sum = sum(TotalSales, na.rm = TRUE), 
    Sales_Mean = mean(TotalSales, na.rm = TRUE), 
    Order_Count = n(), 
    Quantity_Sum = sum(Quantity, na.rm = TRUE), 
    .groups = "drop" 
    ) 

glimpse(sales_summarized)
```


--- 

### 2.2.1 - Bar Chart 

1. Bar graph showing product and the sum of sales for each.

I'm passing **sales_summarized** through a `group_by` function to roll up sales by product. The output is passed to `barplot` for assembling the bar chart. 

```{r}
product_sales <- sales_summarized %>% 
  group_by(Product) %>%
  summarise(Product_Sales = sum(Sales_Sum, na.rm = TRUE), .grops = "DROP") 

barplot( 
  height = product_sales$Product_Sales, 
  names.arg = product_sales$Product, 
  col = "steelblue", 
  main = "Total Sales by Product", 
  xlab = "Product", 
  ylab = "Total Sales"
  )
```

--- 

### 2.2.2 - Pie Chart  

2. Pie chart showing sum of sales by region.

--- 

### 2.2.3 - Box Plot Chart 

3. Box plots for average sales by region and average quantities by region.


--- 

### 2.2.4 - Histogram  

4. Histogram of the count of orders by product.


---

### 2.2.5 - Radial Plot 

5. A radial plot showing total sales (for all products) by month. 


--- 
